---
title: "Macroeconomic dashboard"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.showtext=TRUE, warning = FALSE,
                      message = FALSE, dpi = 96, fig.width = 12, fig.fullwidth=TRUE, fig.height = 5)

library(data.table)
library(here)
library(httr2)
library(czso)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(eurostat)
library(readr)
library(readxl)
library(lubridate)
library(forcats)
library(firatheme)
library(gridExtra)
library(rsdmx)
library(dotenv)
```

```{r fonts}
source("fonts.r")
```


```{r back-2, echo = FALSE, include = FALSE}
# Load functions on start
# Function for lags
  lagp <- function(x, k=1) {
    c(rep(NA, k), x)[1 : length(x)] 
  }
  
# Function for differences
  diffp <- function(x, k=1) {
    lagx <- lagp(x,k)
    x-lagx
  }

# Function for demeaning
  demean <- function(x, na.rm=FALSE) {
    x-mean(x, na.rm=na.rm)
  }
  
# Function for growth rate
  grate <- function(x,k=1) {
    (x-lagp(x,k))/lagp(x,k)
  }

#
#

##########################################################
# span of data, get the current year
date <- Sys.Date()
year_sel <- format(date, "%Y")
start_date <- floor_date(date - months(12), "month")
this_month <- floor_date(date, "month")

#
#

##########################################################
# unemployment
##############
unemp <- get_eurostat('ei_lmhr_m')

unemp_cz <- as.data.table(unemp)
unemp_cz <- unemp_cz[geo == "CZ" & year(TIME_PERIOD) >= (as.numeric(year_sel) - 1) &  s_adj == "SA" & indic == "LM-UN-T-TOT", ]

# select required  columns
unemp_cz <- unemp_cz[, .(time = TIME_PERIOD, value = values, id = 'Eurostat', variable = 'urate')]

#
#
# Podíl nezaměstnaných osob (MPSV - úřady práce)
# 
# primarily: 'https://www.mpsv.cz/documents/20142/1384563/stat-yyyy-mm.zip
# alternatively: 'https://www.mpsv.cz/o/rest/statistiky/nezamestnanost/yyyy/mm'
# 
# celé špatně - již nefční linky MPSV, nově přes otevřená data
# historicky potřeba scrape


# # create the link base on current date
# last_month <- paste0('0', month(date) - 1)
# mpsv_data_link <- paste0('https://www.mpsv.cz/documents/20142/1384563/stat-',year_sel,'-',last_month,'.zip')
# mpsv_file_name <- paste0('4. Nez', last_month, substr(year_sel,3,4),'h.xlsx')

# # try downloading data, if available and contains new value, then update the mpsv_data.csv
# tryCatch({download.file(mpsv_data_link, here('mpsv_data.zip'), mode='wb')
#           unzip('mpsv_data.zip', files = paste0('4. Nez',last_month, substr(year_sel,3,4),'h.xlsx'), exdir=here())
#           u_mpsv <- read_excel(mpsv_file_name, sheet = 'nuts3', range="AD101:AD101", col_names = F)
#           vacancies_mpsv <- read_excel(mpsv_file_name, sheet = 'nuts3', range="X101:X101", col_names = F)
#           unemployed_mpsv <- read_excel(mpsv_file_name, sheet = 'nuts3', range="J101:J101", col_names = F)
               
#           # if available and contains new value, then add it to the data                  
#           mpsv_data_current <- data.table(read_csv('mpsv_data.csv'))
#           # reformat the date
#           mpsv_data_current[, time := as.Date(time, '%d-%m-%y')]

#           unem_mpsv <- data.table(time = as.Date(paste0('01-',last_month,'-',year_sel), '%d-%m-%y'), u_mpsv, vacancies_mpsv, unemployed_mpsv, id = 'MPSV') # id for identification in figures
#           setnames(unem_mpsv, c('time','urate','Vacancies','Unemployed','id'))

#           ifelse(unem_mpsv$time > max(mpsv_data_current$time), assign('mpsv_data', rbind(mpsv_data_current, unem_mpsv)), assign('mpsv_data', mpsv_data_current))
#           write.csv(mpsv_data, 'mpsv_data.csv', row.names = F)
#           },
#           error = function(err){
#           print('The data for last month is not yet available')
#           },
#           warning = function(war){
#           print('The data for last month is not yet available')
#           })

# # read the basic data
# mpsv_data <- data.table(read_csv('mpsv_data.csv'))
# # reformat the date
# mpsv_data[, time := as.Date(time, '%d-%m-%y')]

# # melt the data to long format
# mpsv_data_long <- data.table(melt(mpsv_data, id = c('time','id')))

# #
# #
# # merge the unemployment together
# unemp_final <- rbind(unemp_cz, mpsv_data_long)

# unemployment rates
fig_urate <- ggplot(unemp_cz[variable == 'urate',], aes(x = time, y = value, colour = id, group = id)) +
               xlab('') + ylab('') + geom_line() + geom_point() +
               scale_x_date(labels = scales::label_date_short(format = c('%Y','%b','','')), date_breaks = "months", 
               expand = expansion(c(0.01, 0.01)), limits = c(min(unemp_cz$time), this_month)) + scale_y_continuous(limits = c(1.5,max(unemp_cz$value)+1)) +
               scale_colour_fira(name = element_text("")) + theme_fira() + 
               theme(legend.position = "right", legend.title = element_text(face='bold'),
               legend.margin=margin(0,0,0,0),
               legend.box.margin=margin(-10,-10,-10,-10),
               plot.margin = margin(0,5,10,-10),
               plot.title = element_text(hjust = 0), 
               plot.subtitle = element_text(hjust=0),
               plot.caption = element_text(hjust=0, margin = margin(15,0,-8,0))) + 
               labs(title='Unemployment rate (%)', 
                    caption = paste0('https://ec.europa.eu/eurostat/web/products-datasets/product?code=ei_lmhr_m\n','last checked: ', format(date, '%d-%m-%Y')))

# vacancies and unemployed
# fig_unem <- ggplot(unemp_final[variable %in% c('Vacancies','Unemployed'),], aes(x = time, y = value, 
#                                                                                 colour = variable, 
#                                                                                 group = variable)) +
#                xlab('') + ylab('') +
#                scale_x_date(labels = scales::label_date(format = "%m"), date_breaks = "months", 
#                expand = expansion(c(0.01, 0.01)), limits = c(min(unemp_cz$time), as.Date('2020-12-01'))) +                       scale_y_continuous(labels = scales::label_comma()) +
#                scale_colour_fira(name = element_text("")) + theme_fira() + 
#                theme(legend.position = "bottom", legend.title = element_text(face='bold'), 
#                plot.title =  element_text(hjust = 0), legend.margin=margin(0,0,0,0),
#                legend.box.margin=margin(-10,-10,-10,-10),
#                plot.margin = margin(0,5,10,-5),
#                plot.subtitle = element_text(hjust=0), 
#                plot.caption = element_text(hjust=0, margin = margin(15,0,-8,0))) + 
#                labs(title = 'Unemployment and vacancies', caption = paste0('https://www.mpsv.cz/web/cz/mesicni\n','last checked: ', format(date, '%d-%m-%Y')))

#
#

##
##########################################################
# sales in services
####################
##

services <- data.table(czso_get_table('030030'))
# services_sch <- czso_get_table_schema("030030")

# list of nace categories
cznace <- services[, .(cznace_kod, cznace_txt)]
cznace <- cznace[!duplicated(cznace_kod),]
# cznace # check

# filter the values
# seasonally adjusted, current prices, same period prevous year comparison, all services
# selection of categories: services overall and then subcategories
#
categ <- c("49820001",'45470001','55','56','J','H','68','N','69740001')
categ_txt <- c('Overall (NACE I,J,H,L,M,N)','Retail','Accommodation','Food and beverage','Information and communication',
               'Transportation','Real estate','Administrative','Professional, technical, and research')
categ_concat <- data.table(categ, categ_txt)


serv_current <- services[as_date(paste0(services$mesic, "-", services$rok), format = "%m-%Y") >= start_date & ocisteni_kod == 'O' & is.na(oceneni_kod) &
                      cznace_kod %in% categ_concat$categ & casz_kod == 'C',]
#                
#

# only keep required columns
serv_current <- serv_current[, .(hodnota, mesic, rok, categ = cznace_kod)]

# merge with text value of nace category
serv_current <- serv_current[categ_concat, on = c('categ')]

# reformat month number
serv_current[, mesic := ifelse(nchar(mesic) == 1, paste0('0', mesic), mesic)]

# create a date value
serv_current[, time := as.Date(paste('01',mesic,rok, sep = '-'), '%d-%m-%Y')]

# create split column for overall services and sub-categories
serv_current[categ_txt == 'Overall (NACE I,J,H,L,M,N)', plot := 'Overall (NACE I,J,H,L,M,N)']
serv_current[is.na(plot), plot := 'Subcategories']
serv_current[, categ_factor := factor(categ_txt, levels = c('Overall (NACE I,J,H,L,M,N)',
                                                             'Accommodation',
                                                             'Administrative',
                                                             'Food and beverage',
                                                             'Information and communication',
                                                             'Professional, technical, and research',
                                                             'Real estate',
                                                             'Retail',
                                                             'Transportation'))]

#
#

# plot
fig_sales_serv <- ggplot(serv_current, aes(x = time, y= hodnota, group = categ_factor, colour = categ_factor))

#
#


##
##########################################################
# industry production
#####################
##

industry <- data.table(czso_get_table('150196'))

# filter the specific entries from data, same period previous year index, industry overall, and automotive
industry_sel <- industry[cznace_kod %in% c('05350001','29') & as_date(paste0(industry$mesic, "-", industry$rok), format = "%m-%Y") >= start_date & casz_kod == 'C',]

# only keep required columns
ind_current <- industry_sel[, .(hodnota, mesic, rok, categ = cznace_kod)]

# reformat month number
ind_current[, mesic := ifelse(nchar(mesic) == 1, paste0('0', mesic), mesic)]

# create a date value
ind_current[, time := as.Date(paste('01',mesic,rok, sep = '-'), '%d-%m-%Y')]

ind_current[categ == '29', categ_txt := 'Motor vehicles']
ind_current[categ == '05350001', categ_txt := 'Overall (NACE B,C,D)']

ind_current[, categ_factor := factor(categ_txt, levels = c('Overall (NACE B,C,D)','Motor vehicles'))]

#
#

# plot
fig_prod_ind <- ggplot(ind_current, aes(x = time, y= hodnota, group = categ_factor))

#
#

##
##########################################################
# construction
#################
##

construction <- data.table(czso_get_table('200075'))

# filter the specific entries from dasta, same period previous year index, industry overall, and automotive
construction_sel <- construction[as_date(paste0(construction$mesic, "-", construction$rok), format = "%m-%Y") >= start_date & casz_kod == 'C' & ocisteni_kod == 'P',]

# only keep required columns
cons_current <- construction_sel[, .(hodnota, mesic, rok, stavprace_kod)]

# reformat month number
cons_current[, mesic := ifelse(nchar(mesic) == 1, paste0('0', mesic), mesic)]

# create a date value
cons_current[, time := as.Date(paste('01',mesic,rok, sep = '-'), '%d-%m-%Y')]

cons_current[stavprace_kod == '024', categ_txt := 'Civil engineering']
cons_current[stavprace_kod == '023', categ_txt := 'Construction of buildings']
cons_current[is.na(stavprace_kod), categ_txt := 'Overall (NACE F)']

cons_current[, categ_factor := factor(categ_txt, levels = c('Overall (NACE F)',"Construction of buildings",'Civil engineering'))]
cons_current[!(is.na(stavprace_kod)), plot := 'Subcategories']
cons_current[is.na(stavprace_kod), plot := 'Overall (NACE F)']

#
#

# plot
fig_prod_cons <- ggplot(cons_current, aes(x = time, y= hodnota, group = categ_factor, colour = categ_factor))

#
#

##
##########################################################
# Prices
################
##

# download sdmx data from CZSO
# download CPI index
download.file('https://vdb.czso.cz/pll/eweb/sdmx.getXML?d=CPI', destfile='cpi.xml')

# read into R
cpi_data <- as.data.table(readSDMX('cpi.xml', isURL = F))
cpi_data[, time := as.Date(paste0(TIME_PERIOD,'-01'), '%Y-%m-%d')] # adjust date

# select required columns
cpi_data <- cpi_data[, .(time, value = as.numeric(OBS_VALUE))]

# compute inflation
cpi_data[, value := round(grate(value, k = 12)*100, 1)]

# only select the latest 12 observations
cpi_data <- cpi_data[time >= start_date,]
cpi_data <- cpi_data[, .(time, value, type = "CPI inflation")]

# =============================================================================
# CORE INFLATION
# =============================================================================

# define request
core_infl_req <- request(paste0("https://www.cnb.cz/aradb/api/v1/data?indicator_id_list=SCPIMZM09YOYPECNA&months_before=12&decimal_separator=point&decimals=2&delimiter=comma&api_key=", Sys.getenv('ARAD_API_TOKEN')))

# request
resp <- core_infl_req |> 
  req_perform() |>
  resp_body_string(encoding = 'utf-8')

# transform response data into data table
core_infl_data <- as.data.table(read.table(text = resp, header = T, sep = ','))

# create a corrected datetime
core_infl_data[, time := as.Date(as.character(period), format = '%Y%m%d')] 

# adjust to beginning of the month
core_infl_data[, time := as_date(floor_date(time, "month"))] 

# add series name and drop redundant columns
core_infl_data[, type := 'Core inflation'][
  , c('indicator_id','snapshot_id', 'period') := NULL
]

# bind the CPI and core data together
infl <- rbind(cpi_data, core_infl_data)
infl <- infl[!is.na(value),] # drop NA observations

#
#

# figures
fig_infl <- ggplot(infl, aes(x = time, y = value, group = type, colour = type)) + geom_line() + geom_point() +
               xlab('') + ylab('') +
               scale_x_date(labels = scales::label_date_short(format = c('%Y','%b','','')), date_breaks = "months", 
               expand = expansion(c(0.05, 0.05)), limits = c(min(cpi_data$time), this_month)) +
               scale_y_continuous(labels = scales::label_number(accuracy=0.1)) +
               scale_colour_fira(name = element_text("")) + theme_fira() + 
               theme(legend.position ='bottom', legend.margin=margin(0,0,0,0),
               legend.box.margin=margin(-10,-10,-10,-10),
               plot.margin = margin(0,5,10,-5), legend.title = element_text(face='bold'), 
               plot.title = element_text(hjust = 0), plot.subtitle = element_text(hjust=0), 
               plot.caption = element_text(hjust=0)) +
               labs(title='Inflation', subtitle='Base = same period previous year', caption = paste0('https://vdb.czso.cz/pll/eweb/sdmx.getXML?d=CPI\n', 'https://www.cnb.cz/arad/, Data series SCPIMZM05YOYPECNA\n', ', last checked: ', format(date, '%d-%m-%Y')))

#
#

# download PPI index
download.file('https://vdb.czso.cz/pll/eweb/sdmx.getXML?d=PPI', destfile='ppi.xml')

# read into R
ppi_data <- as.data.table(readSDMX('ppi.xml', isURL = F))
ppi_data[, time := as.Date(paste0(TIME_PERIOD,'-01'), '%Y-%m-%d')] # adjust date

# select required columns
ppi_data <- ppi_data[, .(time, value = as.numeric(OBS_VALUE))]

# compute inflation
ppi_data[, value := round(grate(value, k = 12) * 100,1)]

ppi_data <- ppi_data[time >= start_date, ]

fig_infl_ppi <- ggplot(ppi_data, aes(x = time, y = value)) + geom_line() + geom_point() + xlab('') + ylab('') + 
               scale_x_date(labels = scales::label_date_short(format = c('%Y','%b','','')), date_breaks = "months", 
               expand = expansion(c(0.05, 0.05)), limits = c(min(ppi_data$time), this_month)) +
               scale_y_continuous(labels = scales::label_number(accuracy=0.1)) +
               scale_colour_fira(name = element_text("")) + theme_fira() + 
               theme(legend.position ='bottom', plot.margin = margin(0,5,10,-5), 
               legend.title = element_text(face='bold'), plot.title = element_text(hjust = 0), 
               plot.subtitle = element_text(hjust=0), plot.caption = element_text(hjust=0)) +
               labs(title='PPI inflation', subtitle='Base = same period previous year', caption = paste0( 'https://vdb.czso.cz/pll/eweb/sdmx.getXML?d=PPI\n', 'last checked: ', format(date, '%d-%m-%Y')))

#
#
#

# =============================================================================
# LOANS
# =============================================================================
# source: ARAD CNB
# series SUCM101211XXX801201 - non-performing loans share households
# series SUCM100311XXX801201 - non-performing loans share NFI
# series SMIRSTOBUVMOBJ301CZK1 - HH + NISD loans volume 
# series SMIRSTOBUVMOBJ322CZK1 - HH + NISD consumer loans volume
# series SMIRSTOBUVMOBJ305CZK1 - HH + NISD housing loans volume
# series SMIRNOOBUVMOBJ305CZK013111 - HH + NISD new loans housing
# series SMIRNOOBUVMOBJ304CZK013111 - HH + NISD new consumer loans
# series SMIRSTOBUVMOBJ201CZK1 - NFI loans volume
# series SMIRNOOBUVMOBJ237CZK013111 - NFI new loans CZK
# series SMIRNOOBUVMIRS237CZK013111 - NFI new loans CZK - interest rate

# define request
loans_req <- request(paste0("https://www.cnb.cz/aradb/api/v1/data?indicator_id_list=SMIRSTOBUVMOBJ301CZK1,SMIRSTOBUVMOBJ322CZK1,SMIRSTOBUVMOBJ305CZK1,SMIRSTOBUVMOBJ305CZK1,SMIRNOOBUVMOBJ305CZK013111,SMIRNOOBUVMOBJ304CZK013111,SMIRNOOBUVMOBJ237CZK013111,SMIRNOOBUVMIRS237CZK013111,SMIRSTOBUVMOBJ201CZK1,SUCM100111XXX801201,SUCM100311XXX801201&months_before=25&decimal_separator=point&decimals=2&delimiter=comma&api_key=", Sys.getenv('ARAD_API_KEY')))

# request
resp <- loans_req |> 
  req_perform() |>
  resp_body_string(encoding = 'utf-8')

# transform response data into data table
loans_dt <- as.data.table(read.table(text = resp, header = T, sep = ','))

# create a corrected datetime
loans_dt[, time := as.Date(as.character(period), format = '%Y%m%d')] 

# adjust to beginning of the month and add type column (NFI, HHs)
loans_dt[, time := as_date(floor_date(time, "month"))][
  , type := ifelse(indicator_id %in% c('SMIRSTOBUVMOBJ201CZK1','SUCM100311XXX801201','SMIRNOOBUVMOBJ237CZK013111','SMIRNOOBUVMIRS237CZK013111'), 'Non-financial institutions', 'Households')
] 

# # add series name and drop redundant columns
# loans_dt[, type := 'Core inflation'][
#   , c('indicator_id','snapshot_id', 'period') := NULL
# ]

# transform the data into wide format
loans_dt_wide <- dcast(loans_dt, time + type ~ indicator_id, value.var = 'value')

# compute m-o-m growth rates for loans
loans_dt_wide[, nfi_loans_g := round(grate(SMIRSTOBUVMOBJ201CZK1, k = 12) * 100, 2), by = c('type')]
loans_dt_wide[, hh_loans_g := round(grate(SMIRSTOBUVMOBJ301CZK1, k = 12) * 100, 2), by = c('type')]
loans_dt_wide[, hh_cons_loans_g := round(grate(SMIRSTOBUVMOBJ322CZK1, k = 12) * 100, 2), by = c('type')]
loans_dt_wide[, hh_hous_loans_g := round(grate(SMIRSTOBUVMOBJ305CZK1, k = 12) * 100, 2), by = c('type')]

# drop redundant columns and adjust colnames
loans_dt_wide <- loans_dt_wide[, c('SMIRSTOBUVMOBJ201CZK1','SMIRSTOBUVMOBJ301CZK1','SMIRSTOBUVMOBJ322CZK1','SMIRSTOBUVMOBJ305CZK1') := NULL]
setnames(loans_dt_wide, c('SMIRNOOBUVMIRS237CZK013111','SMIRNOOBUVMOBJ237CZK013111','SMIRNOOBUVMOBJ304CZK013111','SMIRNOOBUVMOBJ305CZK013111','SUCM100111XXX801201','SUCM100311XXX801201'), 
                        c('nfi_new_loans_ir','nfi_new_loans','hh_new_cons_loans','hh_new_hous_loans','npl_nfi','npl_hh'))

# melt data into long format
loans_dt_long <- data.table(melt(loans_dt_wide, id = c('time','type')))

# drop NA values
loans_dt_long <- loans_dt_long[!is.na(value), ]

# take only last 12 periods
loans_dt_long <- loans_dt_long[time >= start_date - months(2), ]

# loans_nfi_data <- data.table(read.table(loans_nfi_link_cnb, header = T, sep='|')) # read the data
# loans_hh_data <- data.table(read.table(loans_hh_link_cnb, header = T, sep='|')) # read the data

# setnames(loans_nfi_data, c('time','loans','npl','npl_share')) # adjust column names
# setnames(loans_hh_data, c('time','loans','npl','npl_share')) # adjust column names

# loans_nfi_data[, type := 'Non-financial institutions']
# loans_hh_data[, type := 'Households']


# loans_nfi_data[, time := as.Date(time, format = '%d.%m.%Y')] # adjust date
# loans_nfi_data[, time := as.Date(paste(year(time),month(time), '01' , sep='-'))] # adjust to beginning of the month for consistency
# loans_hh_data[, time := as.Date(time, format = '%d.%m.%Y')] # adjust date
# loans_hh_data[, time := as.Date(paste(year(time),month(time), '01' , sep='-'))] # adjust to beginning of the month for consistency

# # compute loan growth
# loans_nfi_data <- loans_nfi_data[order(time),]
# loans_nfi_data[, loan_growth := grate(loans, k=1)*100]

# loans_hh_data <- loans_hh_data[order(time),]
# loans_hh_data[, loan_growth := grate(loans, k=1)*100]

# # bind the data
# loans_data <- rbind(loans_hh_data, loans_nfi_data)

# # drop NAs
# loans_data <- loans_data[!is.na(loan_growth),]

# plot loan growth data
fig_loan_growth <- ggplot(loans_dt_long[variable %in% c('nfi_loans_g','hh_loans_g'), ], aes(x = time, y = value, group = type, colour = type)) + geom_line() + geom_point() + xlab('') + ylab('') + 
               scale_x_date(labels = scales::label_date_short(format = c('%Y','%b','','')), date_breaks = "months", 
               expand = expansion(c(0.05, 0.05)), limits = c(min(loans_dt_long$time), this_month)) +
               scale_y_continuous(labels = scales::label_number(accuracy=0.1)) +
               scale_colour_fira(name = element_text("")) + theme_fira() + theme(legend.margin=margin(0,0,0,0),
               legend.box.margin=margin(-10,-10,-10,-10), plot.margin = margin(0,5,10,-5), legend.position ='bottom', legend.title = element_text(face='bold'), plot.title = element_text(hjust = 0), plot.subtitle = element_text(hjust=0), plot.caption = element_text(hjust=0)) +
               labs(title='Loan growth (%)', subtitle='Base = same month previous year', caption = paste0('https://www.cnb.cz/arad/\ndata series: SMIRSTOBUVMOBJ201CZK1, SMIRSTOBUVMOBJ301CZK1\n', 'last checked: ', format(date, '%d-%m-%Y')))

# plot npl data
fig_npls <- ggplot(loans_dt_long[variable %in% c('npl_nfi','npl_hh'), ], aes(x = time, y = value, group = type, colour = type)) + geom_line() + geom_point() + xlab('') + ylab('') + 
               scale_x_date(labels = scales::label_date_short(format = c('%Y','%b','','')), date_breaks = "months", 
               expand = expansion(c(0.05, 0.05)), limits = c(min(loans_dt_long$time), this_month)) +
               scale_y_continuous(labels = scales::label_number(accuracy=0.1)) + # facet_grid(type ~ .) +
               scale_colour_fira(name = element_text("")) + theme_fira() + theme(legend.margin=margin(0,0,0,0),
               legend.box.margin=margin(-10,-10,-10,-10), plot.margin = margin(0,5,10,-5), legend.position ='bottom', legend.title = element_text(face='bold'), plot.title = element_text(hjust = 0), plot.subtitle = element_text(hjust=0), plot.caption = element_text(hjust=0)) +
               labs(title='Non-performing loans', subtitle='% share of total loans', caption = paste0('https://www.cnb.cz/arad/\ndata series: SUCM101211XXX801201, SUCM100311XXX801201\n', 'last checked: ', format(date, '%d-%m-%Y')))

# plot nfi new loans ir
fig_nfi_ir <- ggplot(loans_dt_long[variable %in% c('nfi_new_loans_ir'), ], aes(x = time, y = value, group = type, colour = type)) + geom_line() + geom_point() + xlab('') + ylab('') + 
               scale_x_date(labels = scales::label_date_short(format = c('%Y','%b','','')), date_breaks = "months", 
               expand = expansion(c(0.05, 0.05)), limits = c(min(loans_dt_long$time), this_month)) +
               scale_y_continuous(labels = scales::label_number(accuracy=0.1), limits = c(min(loans_dt_long[variable %in% c('nfi_new_loans_ir'), ]$value)-1, max(loans_dt_long[variable %in% c('nfi_new_loans_ir'), ]$value)+1)) + # facet_grid(type ~ .) +
               scale_colour_fira(name = element_text("")) + theme_fira() + theme(legend.margin=margin(0,0,0,0),
               legend.box.margin=margin(-10,-10,-10,-10), plot.margin = margin(0,5,10,-5), legend.position ='bottom', legend.title = element_text(face='bold'), plot.title = element_text(hjust = 0), plot.subtitle = element_text(hjust=0), plot.caption = element_text(hjust=0)) +
               labs(title='Interest rate on new loans (NFIs)', subtitle='%', caption = paste0('https://www.cnb.cz/arad/\ndata series: SMIRNOOBUVMIRS237CZK013111\n', 'last checked: ', format(date, '%d-%m-%Y')))

# plot nfi new loans ir
fig_nfi_new_loans <- ggplot(loans_dt_long[variable %in% c('nfi_new_loans'), ], aes(x = time, y = value/1000000000, group = type, colour = type)) + geom_line() + geom_point() + xlab('') + ylab('') + 
               scale_x_date(labels = scales::label_date_short(format = c('%Y','%b','','')), date_breaks = "months", 
               expand = expansion(c(0.05, 0.05)), limits = c(min(loans_dt_long$time), this_month)) +
               scale_y_continuous(labels = scales::label_number(accuracy=1), limits = c(min(loans_dt_long[variable %in% c('nfi_new_loans'), ]$value)/10^9-3, max(loans_dt_long[variable %in% c('nfi_new_loans'), ]$value)/10^9+3)) + # facet_grid(type ~ .) +
               scale_colour_fira(name = element_text("")) + theme_fira() + theme(legend.margin=margin(0,0,0,0),
               legend.box.margin=margin(-10,-10,-10,-10), plot.margin = margin(0,5,10,-5), legend.position ='bottom', legend.title = element_text(face='bold'), plot.title = element_text(hjust = 0), plot.subtitle = element_text(hjust=0), plot.caption = element_text(hjust=0)) +
               labs(title='Non-financial institutions new loans (bn. CZK)', subtitle='', caption = paste0('https://www.cnb.cz/arad/\ndata series: SMIRNOOBUVMOBJ237CZK013111\n', 'last checked: ', format(date, '%d-%m-%Y')))

# consumer and housing loans growth y-o-y - households
fig_hh_loan_growth_det <- ggplot(loans_dt_long[variable %in% c('hh_cons_loans_g','hh_hous_loans_g'), ], aes(x = time, y = value, group = variable, colour = variable)) + geom_line() + geom_point() + xlab('') + ylab('') + 
               scale_x_date(labels = scales::label_date_short(format = c("%Y", "%b", "", "")), date_breaks = "months", 
               expand = expansion(c(0.05, 0.05)), limits = c(min(loans_dt_long$time), this_month)) +
               scale_y_continuous(labels = scales::label_number(accuracy=0.1)) +
               scale_colour_fira(name = element_text(""), labels = c('Consumer loans', 'Housing loans')) + theme_fira() + theme(legend.margin=margin(0,0,0,0),
               legend.box.margin=margin(-10,-10,-10,-10), plot.margin = margin(0,5,10,-5), legend.position ='bottom', legend.title = element_text(face='bold'), plot.title = element_text(hjust = 0), plot.subtitle = element_text(hjust=0), plot.caption = element_text(hjust=0)) +
               labs(title='Households loan growth (%)', subtitle='Base = same month previous year', caption = paste0('https://www.cnb.cz/arad/\ndata series: SMIRSTOBUVMOBJ201CZK1, SMIRSTOBUVMOBJ301CZK1\n', 'last checked: ', format(date, '%d-%m-%Y')))

# new consumer and housing loans - households
fig_hh_new_loans_det <- ggplot(loans_dt_long[variable %in% c('hh_new_cons_loans','hh_new_hous_loans'), ], aes(x = time, y = value/10^9, group = variable, colour = variable)) + geom_line() + geom_point() + xlab('') + ylab('') + 
               scale_x_date(labels = scales::label_date_short(format = c("%Y", "%b", "", "")), date_breaks = "months", 
               expand = expansion(c(0.05, 0.05)), limits = c(min(loans_dt_long$time), this_month)) +
               scale_y_continuous(labels = scales::label_number(accuracy=1)) +
               scale_colour_fira(name = element_text(""), labels = c('Consumer loans', 'Housing loans')) + theme_fira() + theme(legend.margin=margin(0,0,0,0),
               legend.box.margin=margin(-10,-10,-10,-10), plot.margin = margin(0,5,10,-5), legend.position ='bottom', legend.title = element_text(face='bold'), plot.title = element_text(hjust = 0), plot.subtitle = element_text(hjust=0), plot.caption = element_text(hjust=0)) +
               labs(title='Households new loans (bn. CZK)', subtitle='', caption = paste0('https://www.cnb.cz/arad/\ndata series: SMIRNOOBUVMOBJ304CZK013111, SMIRNOOBUVMOBJ305CZK013111\n', 'last checked: ', format(date, '%d-%m-%Y')))

#
#

# Loans moratoriums (COVID), data from CNB
# adjust month
# mon <- format(Sys.Date(), format = '%m')
# day <- format(Sys.Date(), format = '%d')

# morat_file_cnb_link <- paste0('https://www.cnb.cz/export/sites/cnb/cs/dohled-financni-trh/.galleries/souhrnne_informace_fin_trhy/statistika_odkladu_splatek_a_uveru_v_programech_covid/tydenni/',as.numeric(year(date)),mon,day,'_statistika_covid.xlsx')

# morat_file_name <- paste0(as.numeric(year(date)),mon,day,'_statistika_covid.xlsx')

# # try downloading data, if available and contains new value, then update the data on moratoriums
# tryCatch({download.file(morat_file_cnb_link, here(morat_file_name), mode='wb')
#           morat_hh <- read_excel(morat_file_name, sheet = 'Moratorium', range="G34:G34", col_names = F)
#           morat_other_hh <- read_excel(morat_file_name, sheet = 'Moratorium', range="G63:G63", col_names = F)
#           morat_nfi <- read_excel(morat_file_name, sheet = 'Moratorium', range="G9:G9", col_names = F)
#           morat_other_nfi <- read_excel(morat_file_name, sheet = 'Moratorium', range="G38:G38", col_names = F)
#           morat_time <- read_excel(morat_file_name, sheet = 'Moratorium', range="B7:B7", col_names = 'time')
          
#           loans_hh <- read_excel(morat_file_name, sheet = 'Moratorium_podily', range="B6:B6", col_names = 'loans_hh')
#           loans_nfi <- read_excel(morat_file_name, sheet = 'Moratorium_podily', range="F6:F6", col_names = 'loans_nfi')
          
#           # if available and contains new value, then add it to the data                  
#           morat_data_current <- data.table(read_csv('morat_data.csv'))
          
#           # reformat the date
#           morat_data_current[, time := as.Date(time, '%d-%m-%y')]
          
#           morat <- data.table(time = as.Date(morat_time$time), morat_hh = morat_hh+morat_other_hh, morat_nfi = morat_nfi + morat_other_nfi, loans_hh*1000, loans_nfi*1000) # id for identification in figures
#           setnames(morat, c('time','morat_hh','morat_nfi','loans_hh','loans_nfi'))
#           morat[, morat_sh_hh := morat_hh/loans_hh]
#           morat[, morat_sh_nfi := morat_nfi/loans_nfi]

#           ifelse(morat$time > max(morat_data_current$time), assign('morat_data', rbind(morat, morat_data_current)), assign('morat_data', morat_data_current))
#           write.csv(morat_data, 'morat_data.csv', row.names = F)
#           },
#           error = function(err){
#           print('The data for the week is not yet available')
#           },
#           warning = function(war){
#           print('The data for the week is not yet available')
#           })

# # adjust the moratorium data
# morat_data <- data.table(read_csv('morat_data.csv'))
# morat_long <- melt(morat_data[, j = .(time,morat_sh_hh,morat_sh_nfi)], id = c('time'), variable.name = 'type', value.name = 'value')
# morat_long[type == 'morat_sh_nfi', type := 'Non-financial institutions']
# morat_long[type == 'morat_sh_hh', type := 'Households']

# # plot loan growth data
# fig_morat <- ggplot(morat_long, aes(x = time, y = value*100, group = type, colour = type)) + geom_line() + geom_point() +              xlab('') + ylab('') + 
#                scale_x_date(labels = scales::label_date(format = "%m"), date_breaks = "months",
#                expand = expansion(c(0.05, 0.05)), limits = c(min(morat_long$time), as.Date('2020-12-01'))) +
#                scale_y_continuous(labels = scales::label_number(accuracy=1)) +
#                scale_colour_fira(name = element_text("")) + theme_fira() + theme(legend.margin=margin(0,0,0,0),
#                legend.box.margin=margin(-10,-10,-10,-10), plot.margin = margin(0,5,10,-5), legend.position ='bottom', legend.title = element_text(face='bold'), plot.title = element_text(hjust = 0), plot.subtitle = element_text(hjust=0), plot.caption = element_text(hjust=0)) +
#                labs(title='Loans in moratorium', subtitle='% share of loans', caption = paste0('https://www.cnb.cz/cs/dohled-financni-trh/souhrnne-informace-fin-trhy/statistika-odkladu-splatek-a-uveru-v-programech-covid/\n','last checked: ', format(date, '%d-%m-%Y')))

# # plot npl data
# fig_npls <- ggplot(loans_data, aes(x = time, y = npl_share, group = type, colour = type)) + geom_line() + geom_point() +                       xlab('') + ylab('') + 
#                scale_x_date(labels = scales::label_date(format = "%m"), date_breaks = "months", 
#                expand = expansion(c(0.05, 0.05)), limits = c(min(loans_data$time), as.Date('2020-12-01'))) +
#                scale_y_continuous(labels = scales::label_number(accuracy=0.1)) +
#                scale_colour_fira(name = element_text("")) + theme_fira() + theme(legend.margin=margin(0,0,0,0),
#                legend.box.margin=margin(-10,-10,-10,-10), plot.margin = margin(0,5,10,-5), legend.position ='bottom', legend.title = element_text(face='bold'), plot.title = element_text(hjust = 0), plot.subtitle = element_text(hjust=0), plot.caption = element_text(hjust=0)) +
#                labs(title='Non-performing loans', subtitle='% share of total loans', caption = paste0('https://www.cnb.cz/cnb/STAT.ARADY_PKG.PARAMETRY_SESTAVY?p_sestuid=22023\n&p_strid=AABBAE&p_lang=CS\n','last checked: ', format(date, '%d-%m-%Y')))
```


<!-- # Macroeconomic dashboard - Czech Republic -->

This dashboard aims to provide basic macroeconomic data for the Czech Republic. The idea is to gather an up-to-date values from different sources and provide an intuitive overview. The initial idea comes from the lack of unified information source about the economic development during the pandemic situation. See also [czechtrack](https://petrbouchal.github.io/czechtrack/) which provides fiscal data and served as inspiration to this endeavour.

# {.tabset .tabset-pills}

## Economy

\ 

```{r, echo = FALSE, out.width='100%', out.height='100%'}
fig_sales_serv + geom_line() + geom_point() + xlab('') + ylab('') + 
               scale_x_date(labels = scales::label_date_short(format = c('%Y','%b','','')), date_breaks = "months", 
               expand = expansion(c(0.05, 0.05)), limits = c(min(serv_current$time), this_month)) +
               scale_y_continuous(breaks = scales::breaks_width(20))+
               scale_colour_fira(name = element_text("")) + theme_fira() + theme(legend.position = "bottom", plot.margin = margin(0,5,10,-5), legend.title = element_text(face='bold'), plot.title = element_text(hjust = 0), plot.subtitle = element_text(hjust=0),
                                                                                 plot.caption = element_text(hjust=0)) +
               facet_wrap(~plot) + labs(title='Sales in services', subtitle='Monthly index, base = same period previous year', caption = paste0('https://vdb.czso.cz/vdbvo2/faces/index.jsf?page=vystup-objekt-parametry&pvo=SLU03-A&sp=A&skupId=1877&pvokc=&katalog=32025&z=T\n','last checked: ', format(date, '%d-%m-%Y')))
```

\

```{r, echo = FALSE, out.width='100%', out.height='100%'}
fig_prod_ind + geom_line() + geom_point() + xlab('') + ylab('') + 
               scale_x_date(labels = scales::label_date_short(format = c('%Y','%b','','')), date_breaks = "months", 
               expand = expansion(c(0.05, 0.05)), limits = c(min(ind_current$time), this_month)) +
               scale_colour_fira(name = element_text("")) + theme_fira() + theme(plot.margin = margin(0,5,10,-5), legend.title = element_text(face='bold'), plot.title = element_text(hjust = 0), plot.subtitle = element_text(hjust=0), plot.caption = element_text(hjust=0)) +
               facet_wrap(~categ_factor) + labs(title='Industry production', subtitle='Monthly index, base = same period previous year', caption = paste0('https://vdb.czso.cz/vdbvo2/faces/index.jsf?page=vystup-objekt-parametry&pvo=PRU01-D&sp=A&skupId=1267&pvokc=&katalog=30835&z=T\n','last checked: ', format(date, '%d-%m-%Y')))
```               

\ 

```{r, echo = FALSE, out.width='100%', out.height='100%'}
fig_prod_cons + geom_line() + geom_point() + xlab('') + ylab('') + 
               scale_x_date(labels = scales::label_date_short(format = c('%Y','%b','','')), date_breaks = "months", 
               expand = expansion(c(0.05, 0.05)), limits = c(min(cons_current$time), this_month)) +
               scale_colour_fira(name = element_text("")) + theme_fira() + theme(legend.position = "bottom", plot.margin = margin(0,5,10,-5), legend.title = element_text(face='bold'), plot.title = element_text(hjust = 0), plot.subtitle = element_text(hjust=0), plot.caption = element_text(hjust=0)) +
               facet_wrap(~plot) + labs(title='Construction production', subtitle='Monthly index, base = same period previous year', caption = paste0('https://vdb.czso.cz/vdbvo2/faces/cs/index.jsf?page=vystup-objekt-parametry&pvo=STA01-D&sp=A&skupId=826&pvokc=&katalog=30836&z=T\n','last checked: ', format(date, '%d-%m-%Y')))
```

\ 

```{r, echo = FALSE, out.width='100%', out.height='100%'}
# plot1 <- fig_urate + geom_line() + geom_point()
# plot2 <- fig_unem + geom_line() + geom_point()
grid.arrange(fig_urate, ncol=2)
```

\ 

```{r, echo = FALSE, out.width='100%', out.height='100%'}
grid.arrange(fig_infl, fig_infl_ppi, ncol=2)
```

\

```{r, echo = FALSE, out.width='100%', out.height='100%'}
grid.arrange(fig_loan_growth, fig_npls, ncol=2)
```

\

```{r, echo = FALSE, out.width='100%', out.height='100%'}
grid.arrange(fig_nfi_new_loans, fig_nfi_ir, ncol=2)
```

\

```{r, echo = FALSE, out.width='100%', out.height='100%'}
grid.arrange(fig_hh_loan_growth_det, fig_hh_new_loans_det, ncol=2)
```

<!-- 
# ```{r, echo = FALSE, out.width='100%', out.height='100%'}
# fig_morat
# ```

## Leading indicators

## Health

# ```{r mort-data}
# zmr0 <- czso_get_table("130185", force_redownload = T)
# zmr0$vek_txt[zmr0$vek_txt == '65-74'] <- '65-84'
# zmr0$vek_txt[zmr0$vek_txt == '75-84'] <- '65-84'

# zmr <- zmr0 %>% 
#   mutate(tyden = as.numeric(tyden),
#          casref_do = as.Date(casref_do),
#          day_sameyear = make_date(1970, month = month(casref_do),
#                                   day = day(casref_do))) %>% 
#   mutate(vek_txt = fct_relevel(vek_txt, "celkem") %>% 
#            fct_recode(Total = "celkem"))

# last_date_fmt <- max(zmr$casref_do)
# today_fmt <- today()
# ```

# ```{r time-adjusting-load}
# vek <- read_rds("vek.rds")
# vek_ciselnik <- read_rds("vek_ciselnik.rds")
# vek_prevodnik <- read_csv("vek-prevodnik.csv")
# ```


# ```{r time-adjusting-age}
# vek_slct <- vek %>%  
#   filter(vuzemi_cis == "97") %>% 
#   filter(is.na(pohlavi_txt)) %>% 
#   left_join(vek_ciselnik %>%
#                     select(vek_kod = CHODNOTA, zkr = ZKRTEXT, 
#                            txt = TEXT, matches("MIN|MAX"))) %>% 
#   select(idhod, hodnota, vek_txt, zkr, txt, casref_do) %>% 
#   left_join(vek_prevodnik) %>% 
#   group_by(casref_do, vek_txt = vek_grp) %>% 
#   summarise(pocet_obyv = sum(hodnota, na.rm = T)) %>% 
#   mutate(rok = year(casref_do)) %>% 
#   replace_na(list(vek_txt = "Total")) %>% 
#   ungroup() %>% 
#   select(-casref_do)

# vek_2019 <- vek_slct %>% 
#   filter(rok == 2019, !is.na(vek_txt)) %>% 
#   rename(obyv_2019 = pocet_obyv) %>% 
#   select(-rok)

# vek_do2020 <- vek_slct %>% 
#   bind_rows(vek_slct %>% 
#               filter(rok == 2019) %>% 
#               mutate(rok = 2020))
# ```


# ```{r time-adjusting-age-adjust}
# zmr_normd <- zmr %>% 
#   filter(vek_txt != "Total") %>% 
#   select(hodnota, rok, tyden, roktyden, vek_txt,
#          idhod, stapro_kod, casref_do) %>% 
#   left_join(vek_do2020) %>% 
#   mutate(per_100k = hodnota/pocet_obyv * 1e5) %>% 
#   left_join(vek_2019) %>% 
#   mutate(normd = per_100k * obyv_2019 / 1e5)

# zmr_normd_with_sum <- zmr_normd %>% 
#   bind_rows(zmr_normd %>% 
#               group_by(rok, tyden, roktyden, casref_do) %>% 
#               summarise(across(c(hodnota, obyv_2019, pocet_obyv, normd), sum)) %>% 
#               mutate(per_100k = hodnota/pocet_obyv * 1e5,
#                      vek_txt = "Total")) %>% 
#   mutate(day_sameyear = make_date(1970, month(casref_do), day(casref_do)))
# ```


# ```{r time-adjusting-prep}
# zmr_ribbon_data <- zmr_normd_with_sum %>% 
#   mutate(is2020 = rok == 2020) %>% 
#   mutate(day_sameyear_floored = floor_date(day_sameyear, "weeks")) %>% 
#   group_by(is2020, vek_txt, tyden, day_sameyear_floored) %>% 
#   summarise(min = min(normd, na.rm = T), 
#             max = max(normd, na.rm = T), .groups = "drop") %>% 
#   mutate(vek_txt = fct_relevel(vek_txt, "Total") %>% 
#            fct_recode(`85+` = "85 a více"))
# ```


# ```{r mort-labels}
# lbs_grey <- tribble(~vek_txt, ~day_sameyear_floored, ~min, ~txt,
#                factor("Total"), as.Date("1970-02-01"), 2650, "range in \n2011-19",
#                )
# lbs_blue <- tribble(~vek_txt, ~day_sameyear_floored, ~min, ~txt,
#                factor("Total"), as.Date("1970-05-05"), 1900, "2020"
#                )
# ```


# ```{r age-adjusted-plot}
# ggplot(data = zmr_ribbon_data, mapping = aes(group = vek_txt)) +
#   geom_ribbon(data = zmr_ribbon_data %>% filter(!is2020),
#               alpha = 0.3,
#               aes(x = day_sameyear_floored, ymin = min, ymax = max)) +
#   geom_line(data = zmr_ribbon_data %>% filter(is2020),
#             colour = "darkblue",
#             aes(x = day_sameyear_floored, y = min)) +
#   geom_point(data = zmr_ribbon_data %>% filter(is2020),
#             aes(x = day_sameyear_floored, y = min), 
#             size = 1, colour = "darkblue", fill = "white", shape = 21) +
#   facet_wrap(~vek_txt, scales = "free_y") +
#   scale_x_date(date_breaks = "3 months",
#                date_labels = "%b", breaks = c(1,3,6, 9, 12)) +
#   theme_fira() +
#   theme(plot.margin = margin(0,5,10,0), 
#         legend.title = element_text(face='bold'), 
#         plot.title = element_text(hjust = 0), 
#         plot.subtitle = element_text(hjust=0),
#         plot.caption = element_text(hjust=0)) +
#   labs(x = NULL, y = NULL,
#        title = "Mortality (all-cause)",
#        subtitle = "Deaths per week, 2011-2020. Blue = 2020, grey = range in 2011-2019",
#        caption = str_glue("Based on CZSO dataset 130185 (deaths) and 130142 (age) | last checked {today_fmt} | Last data for {last_date_fmt}.")) +
#   geom_text(data = lbs_grey, aes(x = day_sameyear_floored, 
#                             y = min, label = txt), 
#             family = "Fira Sans",
#             fontface = "bold",
#             colour = "#454545",size = 3) +
#   geom_text(data = lbs_blue, aes(x = day_sameyear_floored, 
#                             y = min, label = txt), 
#             family = "Fira Sans",
#             fontface = "bold",
#             colour = "darkblue", size = 3)
# ```

## Energy consumption 
#
## PX index
# -->

<!-- ## Some other useful indicators? -->