---
title: "Macroeconomic dashboard"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.showtext=TRUE, warning = FALSE,
                      message = FALSE, dpi = 96, fig.width = 12, fig.fullwidth=TRUE, fig.height = 5)

library(data.table)
library(here)
library(czso)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(eurostat)
library(readr)
library(readxl)
library(lubridate)
library(forcats)
library(firatheme)
library(gridExtra)
library(rsdmx)
```

```{r fonts}
source("fonts.r")
```


```{r back-2, echo = FALSE, include = FALSE}
# Load functions on start
# Function for lags
  lagp <- function(x, k=1) {
    c(rep(NA, k), x)[1 : length(x)] 
  }
  
# Function for differences
  diffp <- function(x, k=1) {
    lagx <- lagp(x,k)
    x-lagx
  }

# Function for demeaning
  demean <- function(x, na.rm=FALSE) {
    x-mean(x, na.rm=na.rm)
  }
  
# Function for growth rate
  grate <- function(x,k=1) {
    (x-lagp(x,k))/lagp(x,k)
  }

#
#

##########################################################
# span of data, get the current year
date <- Sys.Date()
year_sel <- format(date, "%Y")

#
#

##########################################################
# unemployment
##############
unemp <- get_eurostat('ei_lmhr_m')

unemp_cz <- as.data.table(unemp)
unemp_cz <- unemp_cz[geo == "CZ" & year(time) >= (as.numeric(year_sel)-1) &  s_adj == "SA" & indic == "LM-UN-T-TOT", ]

# select required  columns
unemp_cz <- unemp_cz[, .(time, value = values, id = 'Eurostat', variable = 'urate')]

#
#
# Podíl nezaměstnaných osob (MPSV - úřady práce)
# 
# primarily: 'https://www.mpsv.cz/documents/20142/1384563/stat-yyyy-mm.zip
# alternatively: 'https://www.mpsv.cz/o/rest/statistiky/nezamestnanost/yyyy/mm'
# 

# create the link base on current date
last_month <- paste0('0',month(date)-1)
mpsv_data_link <- paste0('https://www.mpsv.cz/documents/20142/1384563/stat-',year_sel,'-',last_month,'.zip')
mpsv_file_name <- paste0('4. Nez',last_month, substr(year_sel,3,4),'h.xlsx')

# try downloading data, if available and contains new value, then update the mpsv_data.csv
tryCatch({download.file(mpsv_data_link, here('mpsv_data.zip'), mode='wb')
          unzip('mpsv_data.zip', files = paste0('4. Nez',last_month, substr(year_sel,3,4),'h.xlsx'), exdir=here())
          u_mpsv <- read_excel(mpsv_file_name, sheet = 'nuts3', range="AD101:AD101", col_names = F)
          vacancies_mpsv <- read_excel(mpsv_file_name, sheet = 'nuts3', range="X101:X101", col_names = F)
          unemployed_mpsv <- read_excel(mpsv_file_name, sheet = 'nuts3', range="J101:J101", col_names = F)
               
          # if available and contains new value, then add it to the data                  
          mpsv_data_current <- data.table(read_csv('mpsv_data.csv'))
          # reformat the date
          mpsv_data_current[, time := as.Date(time, '%d-%m-%y')]

          unem_mpsv <- data.table(time = as.Date(paste0('01-',last_month,'-',year_sel), '%d-%m-%y'), u_mpsv, vacancies_mpsv, unemployed_mpsv, id = 'MPSV') # id for identification in figures
          setnames(unem_mpsv, c('time','urate','Vacancies','Unemployed','id'))

          ifelse(unem_mpsv$time > max(mpsv_data_current$time), assign('mpsv_data', rbind(mpsv_data_current, unem_mpsv)), assign('mpsv_data', mpsv_data_current))
          write.csv(mpsv_data, 'mpsv_data.csv', row.names = F)
          },
          error = function(err){
          print('The data for last month is not yet available')
          },
          warning = function(war){
          print('The data for last month is not yet available')
          })

# read the basic data
mpsv_data <- data.table(read_csv('mpsv_data.csv'))
# reformat the date
mpsv_data[, time := as.Date(time, '%d-%m-%y')]

# melt the data to long format
mpsv_data_long <- data.table(melt(mpsv_data, id = c('time','id')))

#
#
# merge the unemployment together
unemp_final <- rbind(unemp_cz, mpsv_data_long)

# unemployment rates
fig_urate <- ggplot(unemp_final[variable == 'urate',], aes(x = time, y = value, colour = id, group = id)) +
               xlab('') + ylab('') +
               scale_x_date(labels = scales::label_date(format = "%m"), date_breaks = "months", 
               expand = expansion(c(0.01, 0.01)), limits = c(min(unemp_cz$time), as.Date('2020-12-01'))) +                       scale_y_continuous(limits = c(1.5,max(mpsv_data$urate)+1)) +
               scale_colour_fira(name = element_text("")) + theme_fira() + 
               theme(legend.position = "bottom", legend.title = element_text(face='bold'),
               legend.margin=margin(0,0,0,0),
               legend.box.margin=margin(-10,-10,-10,-10),
               plot.margin = margin(0,5,10,-10),
               plot.title = element_text(hjust = 0), 
               plot.subtitle = element_text(hjust=0),
               plot.caption = element_text(hjust=0, margin = margin(15,0,-8,0))) + 
               labs(title='Unemployment rate (%)', 
                    caption = paste0('https://ec.europa.eu/eurostat/web/products-datasets/product?code=ei_lmhr_m\nhttps://www.mpsv.cz/web/cz/mesicni\n','last checked: ', format(date, '%d-%m-%Y')))

# vacancies and unemployed
fig_unem <- ggplot(unemp_final[variable %in% c('Vacancies','Unemployed'),], aes(x = time, y = value, 
                                                                                colour = variable, 
                                                                                group = variable)) +
               xlab('') + ylab('') +
               scale_x_date(labels = scales::label_date(format = "%m"), date_breaks = "months", 
               expand = expansion(c(0.01, 0.01)), limits = c(min(unemp_cz$time), as.Date('2020-12-01'))) +                       scale_y_continuous(labels = scales::label_comma()) +
               scale_colour_fira(name = element_text("")) + theme_fira() + 
               theme(legend.position = "bottom", legend.title = element_text(face='bold'), 
               plot.title =  element_text(hjust = 0), legend.margin=margin(0,0,0,0),
               legend.box.margin=margin(-10,-10,-10,-10),
               plot.margin = margin(0,5,10,-5),
               plot.subtitle = element_text(hjust=0), 
               plot.caption = element_text(hjust=0, margin = margin(15,0,-8,0))) + 
               labs(title = 'Unemployment and vacancies', caption = paste0('https://www.mpsv.cz/web/cz/mesicni\n','last checked: ', format(date, '%d-%m-%Y')))

#
#

##
##########################################################
# sales in services
####################
##

services <- data.table(czso_get_table('030030'))
# services_sch <- czso_get_table_schema("030030")

# list of nace categories
cznace <- services[, .(cznace_kod, cznace_txt)]
cznace <- cznace[!duplicated(cznace_kod),]
# cznace # check

# filter the values
# seasonally adjusted, current prices, same period prevous year comparison, all services
# selection of categories: services overall and then subcategories
#
categ <- c("49820001",'45470001','55','56','J','H','68','N','69740001')
categ_txt <- c('Overall (NACE I,J,H,L,M,N)','Retail','Accommodation','Food and beverage','Information and communication',
               'Transportation','Real estate','Administrative','Professional, technical, and research')
categ_concat <- data.table(categ, categ_txt)

serv_current <- services[rok >= as.numeric(year_sel) & ocisteni_kod == 'O' & is.na(oceneni_kod) &
                      cznace_kod %in% categ_concat$categ & casz_kod == 'C',]

#                
#

# only keep required columns
serv_current <- serv_current[, .(hodnota, mesic, rok, categ = cznace_kod)]

# merge with text value of nace category
serv_current <- serv_current[categ_concat, on = c('categ')]

# reformat month number
serv_current[, mesic := ifelse(nchar(mesic) == 1, paste0('0', mesic), mesic)]

# create a date value
serv_current[, time := as.Date(paste('01',mesic,rok, sep = '-'), '%d-%m-%Y')]

# create split column for overall services and sub-categories
serv_current[categ_txt == 'Overall (NACE I,J,H,L,M,N)', plot := 'Overall (NACE I,J,H,L,M,N)']
serv_current[is.na(plot), plot := 'Subcategories']
serv_current[, categ_factor := factor(categ_txt, levels = c('Overall (NACE I,J,H,L,M,N)',
                                                             'Accommodation',
                                                             'Administrative',
                                                             'Food and beverage',
                                                             'Information and communication',
                                                             'Professional, technical, and research',
                                                             'Real estate',
                                                             'Retail',
                                                             'Transportation'))]

#
#

# plot
fig_sales_serv <- ggplot(serv_current, aes(x = time, y= hodnota, group = categ_factor, colour = categ_factor))

#
#


##
##########################################################
# industry production
#####################
##

industry <- data.table(czso_get_table('150196'))

# filter the specific entries from data, same period previous year index, industry overall, and automotive
industry_sel <- industry[cznace_kod %in% c('05350001','29') & rok == as.numeric(year_sel) & casz_kod == 'C',]

# only keep required columns
ind_current <- industry_sel[, .(hodnota, mesic, rok, categ = cznace_kod)]

# reformat month number
ind_current[, mesic := ifelse(nchar(mesic) == 1, paste0('0', mesic), mesic)]

# create a date value
ind_current[, time := as.Date(paste('01',mesic,rok, sep = '-'), '%d-%m-%Y')]

ind_current[categ == '29', categ_txt := 'Motor vehicles']
ind_current[categ == '05350001', categ_txt := 'Overall (NACE B,C,D)']

ind_current[, categ_factor := factor(categ_txt, levels = c('Overall (NACE B,C,D)','Motor vehicles'))]

#
#

# plot
fig_prod_ind <- ggplot(ind_current, aes(x = time, y= hodnota, group = categ_factor))

#
#

##
##########################################################
# construction
#################
##

construction <- data.table(czso_get_table('200075'))

# filter the specific entries from dasta, same period previous year index, industry overall, and automotive
construction_sel <- construction[rok == as.numeric(year_sel) & casz_kod == 'C' & ocisteni_kod == 'P',]

# only keep required columns
cons_current <- construction_sel[, .(hodnota, mesic, rok, stavprace_kod)]

# reformat month number
cons_current[, mesic := ifelse(nchar(mesic) == 1, paste0('0', mesic), mesic)]

# create a date value
cons_current[, time := as.Date(paste('01',mesic,rok, sep = '-'), '%d-%m-%Y')]

cons_current[stavprace_kod == '024', categ_txt := 'Civil engineering']
cons_current[stavprace_kod == '023', categ_txt := 'Construction of buildings']
cons_current[is.na(stavprace_kod), categ_txt := 'Overall (NACE F)']

cons_current[, categ_factor := factor(categ_txt, levels = c('Overall (NACE F)',"Construction of buildings",'Civil engineering'))]
cons_current[!(is.na(stavprace_kod)), plot := 'Subcategories']
cons_current[is.na(stavprace_kod), plot := 'Overall (NACE F)']

#
#

# plot
fig_prod_cons <- ggplot(cons_current, aes(x = time, y= hodnota, group = categ_factor, colour = categ_factor))

#
#

##
##########################################################
# Prices
################
##

# download sdmx data from CZSO
# download CPI index
download.file('https://vdb.czso.cz/pll/eweb/sdmx.getXML?d=CPI', destfile='cpi.xml')

# read into R
cpi_data <- as.data.table(readSDMX('cpi.xml', isURL = F))
cpi_data[, time := as.Date(paste0(TIME_PERIOD,'-01'), '%Y-%m-%d')] # adjust date

# select required columns
cpi_data <- cpi_data[, .(time, value = as.numeric(OBS_VALUE))]

# compute inflation
cpi_data[, value := round(grate(value, k = 12)*100,1)]

cpi_data <- cpi_data[year(time) == year_sel,]
cpi_data <- cpi_data[,.(time, value, type="CPI inflation")]

#
#

# Core inflation
# adjust month
mon <- ifelse(nchar(month(date)) == 1, paste0('0', month(date)), month(date))

# create link for cnb data
core_link_cnb <- paste0('https://www.cnb.cz/cnb/STAT.ARADY_PKG.VYSTUP?p_period=1&p_sort=2&p_des=50&p_sestuid=21727&p_uka=9&p_strid=ACBAA&p_od=',year(date),'01','&p_do=',year(date),mon,'&p_lang=CS&p_format=2&p_decsep=.')

core_infl_data <- data.table(read.table(core_link_cnb, header = T, sep='|')) # read the data

setnames(core_infl_data, c('time','value')) # adjust column names
# adjust date
core_infl_data[, time := as.Date(time, format = '%d.%m.%Y')] # adjust date
core_infl_data[, time := as.Date(paste(year(time),month(time), '01' , sep='-'))] # adjust to beginning of the month for consistency
core_infl_data[, type := 'Core inflation']

# bind the CPI and core data together
infl <- rbind(cpi_data, core_infl_data)
infl <- infl[!is.na(value),] # drop NA observations

#
#

# figures
fig_infl <- ggplot(infl, aes(x = time, y = value, group = type, colour = type)) + geom_line() + geom_point() +
               xlab('') + ylab('') +
               scale_x_date(labels = scales::label_date(format = "%m"), date_breaks = "months", 
               expand = expansion(c(0.05, 0.05)), limits = c(min(cpi_data$time), as.Date('2020-12-01'))) +
               scale_y_continuous(labels = scales::label_number(accuracy=0.1)) +
               scale_colour_fira(name = element_text("")) + theme_fira() + 
               theme(legend.position ='bottom', legend.margin=margin(0,0,0,0),
               legend.box.margin=margin(-10,-10,-10,-10),
               plot.margin = margin(0,5,10,-5), legend.title = element_text(face='bold'), 
               plot.title = element_text(hjust = 0), plot.subtitle = element_text(hjust=0), 
               plot.caption = element_text(hjust=0)) +
               labs(title='Inflation', subtitle='Base = same period previous year', caption = paste0('https://vdb.czso.cz/pll/eweb/sdmx.getXML?d=CPI\n','https://www.cnb.cz/cnb/STAT.ARADY_PKG.PARAMETRY_SESTAVY?p_sestuid=21727&\np_strid=ACBAA&p_lang=CS\n','last checked: ', format(date, '%d-%m-%Y')))

#
#

# download PPI index
download.file('https://vdb.czso.cz/pll/eweb/sdmx.getXML?d=PPI', destfile='ppi.xml')

# read into R
ppi_data <- as.data.table(readSDMX('ppi.xml', isURL = F))
ppi_data[, time := as.Date(paste0(TIME_PERIOD,'-01'), '%Y-%m-%d')] # adjust date

# select required columns
ppi_data <- ppi_data[, .(time, value = as.numeric(OBS_VALUE))]

# compute inflation
ppi_data[, value := round(grate(value, k = 12)*100,1)]

ppi_data <- ppi_data[year(time) == year_sel,]

fig_infl_ppi <- ggplot(ppi_data, aes(x = time, y = value)) + geom_line() + geom_point() + xlab('') + ylab('') + 
               scale_x_date(labels = scales::label_date(format = "%m"), date_breaks = "months", 
               expand = expansion(c(0.05, 0.05)), limits = c(min(ppi_data$time), as.Date('2020-12-01'))) +
               scale_y_continuous(labels = scales::label_number(accuracy=0.1)) +
               scale_colour_fira(name = element_text("")) + theme_fira() + 
               theme(legend.position ='bottom', plot.margin = margin(0,5,10,-5), 
               legend.title = element_text(face='bold'), plot.title = element_text(hjust = 0), 
               plot.subtitle = element_text(hjust=0), plot.caption = element_text(hjust=0)) +
               labs(title='PPI inflation', subtitle='Base = same period previous year', caption = paste0( 'https://vdb.czso.cz/pll/eweb/sdmx.getXML?d=PPI\n', 'last checked: ', format(date, '%d-%m-%Y')))

#
#
#

# Loans from CNB
# adjust month
mon <- ifelse(nchar(month(date)) == 1, paste0('0', month(date)), month(date))

loans_nfi_link_cnb <- paste0('https://www.cnb.cz/cnb/STAT.ARADY_PKG.VYSTUP?p_period=1&p_sort=2&p_des=50&p_sestuid=22023&p_uka=7%2C8%2C9&p_strid=AABBAE&p_od=',as.numeric(year(date))-1,'12','&p_do=',year(date),mon,'&p_lang=CS&p_format=2&p_decsep=.')
loans_hh_link_cnb <- paste0('https://www.cnb.cz/cnb/STAT.ARADY_PKG.VYSTUP?p_period=1&p_sort=2&p_des=50&p_sestuid=22023&p_uka=16%2C17%2C18&p_strid=AABBAE&p_od=',as.numeric(year(date))-1,'12','&p_do=',year(date),mon,'&p_lang=CS&p_format=2&p_decsep=.')


loans_nfi_data <- data.table(read.table(loans_nfi_link_cnb, header = T, sep='|')) # read the data
loans_hh_data <- data.table(read.table(loans_hh_link_cnb, header = T, sep='|')) # read the data

setnames(loans_nfi_data, c('time','loans','npl','npl_share')) # adjust column names
setnames(loans_hh_data, c('time','loans','npl','npl_share')) # adjust column names

loans_nfi_data[, type := 'Non-financial institutions']
loans_hh_data[, type := 'Households']


loans_nfi_data[, time := as.Date(time, format = '%d.%m.%Y')] # adjust date
loans_nfi_data[, time := as.Date(paste(year(time),month(time), '01' , sep='-'))] # adjust to beginning of the month for consistency
loans_hh_data[, time := as.Date(time, format = '%d.%m.%Y')] # adjust date
loans_hh_data[, time := as.Date(paste(year(time),month(time), '01' , sep='-'))] # adjust to beginning of the month for consistency

# compute loan growth
loans_nfi_data <- loans_nfi_data[order(time),]
loans_nfi_data[, loan_growth := grate(loans, k=1)*100]

loans_hh_data <- loans_hh_data[order(time),]
loans_hh_data[, loan_growth := grate(loans, k=1)*100]

# bind the data
loans_data <- rbind(loans_hh_data, loans_nfi_data)

# drop NAs
loans_data <- loans_data[!is.na(loan_growth),]

# plot loan growth data
fig_loan_growth <- ggplot(loans_data, aes(x = time, y = loan_growth, group = type, colour = type)) + geom_line() + geom_point() +              xlab('') + ylab('') + 
               scale_x_date(labels = scales::label_date(format = "%m"), date_breaks = "months", 
               expand = expansion(c(0.05, 0.05)), limits = c(min(loans_data$time), as.Date('2020-12-01'))) +
               scale_y_continuous(labels = scales::label_number(accuracy=0.1)) +
               scale_colour_fira(name = element_text("")) + theme_fira() + theme(legend.margin=margin(0,0,0,0),
               legend.box.margin=margin(-10,-10,-10,-10), plot.margin = margin(0,5,10,-5), legend.position ='bottom', legend.title = element_text(face='bold'), plot.title = element_text(hjust = 0), plot.subtitle = element_text(hjust=0), plot.caption = element_text(hjust=0)) +
               labs(title='Loan growth (%)', subtitle='Base = previous month', caption = paste0('https://www.cnb.cz/cnb/STAT.ARADY_PKG.PARAMETRY_SESTAVY?p_sestuid=22023\n&p_strid=AABBAE&p_lang=CS\n','last checked: ', format(date, '%d-%m-%Y')))

# plot npl data
fig_npls <- ggplot(loans_data, aes(x = time, y = npl_share, group = type, colour = type)) + geom_line() + geom_point() +                       xlab('') + ylab('') + 
               scale_x_date(labels = scales::label_date(format = "%m"), date_breaks = "months", 
               expand = expansion(c(0.05, 0.05)), limits = c(min(loans_data$time), as.Date('2020-12-01'))) +
               scale_y_continuous(labels = scales::label_number(accuracy=0.1)) +
               scale_colour_fira(name = element_text("")) + theme_fira() + theme(legend.margin=margin(0,0,0,0),
               legend.box.margin=margin(-10,-10,-10,-10), plot.margin = margin(0,5,10,-5), legend.position ='bottom', legend.title = element_text(face='bold'), plot.title = element_text(hjust = 0), plot.subtitle = element_text(hjust=0), plot.caption = element_text(hjust=0)) +
               labs(title='Non-performing loans', subtitle='% share of total loans', caption = paste0('https://www.cnb.cz/cnb/STAT.ARADY_PKG.PARAMETRY_SESTAVY?p_sestuid=22023\n&p_strid=AABBAE&p_lang=CS\n','last checked: ', format(date, '%d-%m-%Y')))

#
#

# Loans moratoriums (COVID), data from CNB
# adjust month
mon <- format(Sys.Date(), format = '%m')
day <- format(Sys.Date(), format = '%d')

morat_file_cnb_link <- paste0('https://www.cnb.cz/export/sites/cnb/cs/dohled-financni-trh/.galleries/souhrnne_informace_fin_trhy/statistika_odkladu_splatek_a_uveru_v_programech_covid/tydenni/',as.numeric(year(date)),mon,day,'_statistika_covid.xlsx')

morat_file_name <- paste0(as.numeric(year(date)),mon,day,'_statistika_covid.xlsx')

# try downloading data, if available and contains new value, then update the data on moratoriums
tryCatch({download.file(morat_file_cnb_link, here(morat_file_name), mode='wb')
          morat_hh <- read_excel(morat_file_name, sheet = 'Moratorium', range="G34:G34", col_names = F)
          morat_other_hh <- read_excel(morat_file_name, sheet = 'Moratorium', range="G63:G63", col_names = F)
          morat_nfi <- read_excel(morat_file_name, sheet = 'Moratorium', range="G9:G9", col_names = F)
          morat_other_nfi <- read_excel(morat_file_name, sheet = 'Moratorium', range="G38:G38", col_names = F)
          morat_time <- read_excel(morat_file_name, sheet = 'Moratorium', range="B7:B7", col_names = 'time')
          
          loans_hh <- read_excel(morat_file_name, sheet = 'Moratorium_podily', range="B6:B6", col_names = 'loans_hh')
          loans_nfi <- read_excel(morat_file_name, sheet = 'Moratorium_podily', range="F6:F6", col_names = 'loans_nfi')
          
          # if available and contains new value, then add it to the data                  
          morat_data_current <- data.table(read_csv('morat_data.csv'))
          
          # reformat the date
          morat_data_current[, time := as.Date(time, '%d-%m-%y')]
          
          morat <- data.table(time = as.Date(morat_time$time), morat_hh = morat_hh+morat_other_hh, morat_nfi = morat_nfi + morat_other_nfi, loans_hh*1000, loans_nfi*1000) # id for identification in figures
          setnames(morat, c('time','morat_hh','morat_nfi','loans_hh','loans_nfi'))
          morat[, morat_sh_hh := morat_hh/loans_hh]
          morat[, morat_sh_nfi := morat_nfi/loans_nfi]

          ifelse(morat$time > max(morat_data_current$time), assign('morat_data', rbind(morat, morat_data_current)), assign('morat_data', morat_data_current))
          write.csv(morat_data, 'morat_data.csv', row.names = F)
          },
          error = function(err){
          print('The data for the week is not yet available')
          },
          warning = function(war){
          print('The data for the week is not yet available')
          })

# adjust the moratorium data
morat_data <- data.table(read_csv('morat_data.csv'))
morat_long <- melt(morat_data[, j = .(time,morat_sh_hh,morat_sh_nfi)], id = c('time'), variable.name = 'type', value.name = 'value')
morat_long[type == 'morat_sh_nfi', type := 'Non-financial institutions']
morat_long[type == 'morat_sh_hh', type := 'Households']

# plot loan growth data
fig_morat <- ggplot(morat_long, aes(x = time, y = value*100, group = type, colour = type)) + geom_line() + geom_point() +              xlab('') + ylab('') + 
               scale_x_date(labels = scales::label_date(format = "%m"), date_breaks = "months",
               expand = expansion(c(0.05, 0.05)), limits = c(min(morat_long$time), as.Date('2020-12-01'))) +
               scale_y_continuous(labels = scales::label_number(accuracy=1)) +
               scale_colour_fira(name = element_text("")) + theme_fira() + theme(legend.margin=margin(0,0,0,0),
               legend.box.margin=margin(-10,-10,-10,-10), plot.margin = margin(0,5,10,-5), legend.position ='bottom', legend.title = element_text(face='bold'), plot.title = element_text(hjust = 0), plot.subtitle = element_text(hjust=0), plot.caption = element_text(hjust=0)) +
               labs(title='Loans in moratorium', subtitle='% share of loans', caption = paste0('https://www.cnb.cz/cs/dohled-financni-trh/souhrnne-informace-fin-trhy/statistika-odkladu-splatek-a-uveru-v-programech-covid/\n','last checked: ', format(date, '%d-%m-%Y')))

# plot npl data
fig_npls <- ggplot(loans_data, aes(x = time, y = npl_share, group = type, colour = type)) + geom_line() + geom_point() +                       xlab('') + ylab('') + 
               scale_x_date(labels = scales::label_date(format = "%m"), date_breaks = "months", 
               expand = expansion(c(0.05, 0.05)), limits = c(min(loans_data$time), as.Date('2020-12-01'))) +
               scale_y_continuous(labels = scales::label_number(accuracy=0.1)) +
               scale_colour_fira(name = element_text("")) + theme_fira() + theme(legend.margin=margin(0,0,0,0),
               legend.box.margin=margin(-10,-10,-10,-10), plot.margin = margin(0,5,10,-5), legend.position ='bottom', legend.title = element_text(face='bold'), plot.title = element_text(hjust = 0), plot.subtitle = element_text(hjust=0), plot.caption = element_text(hjust=0)) +
               labs(title='Non-performing loans', subtitle='% share of total loans', caption = paste0('https://www.cnb.cz/cnb/STAT.ARADY_PKG.PARAMETRY_SESTAVY?p_sestuid=22023\n&p_strid=AABBAE&p_lang=CS\n','last checked: ', format(date, '%d-%m-%Y')))
```

<!-- # Macroeconomic dashboard - Czech Republic -->

This dashboard aims to provide basic macroeconomic data for the Czech Republic. The idea is to gather an up-to-date values from different sources and provide an intuitive overview. The initial idea comes from the lack of unified information source about the economic development during the pandemic situation. See also [czechtrack](https://petrbouchal.github.io/czechtrack/) which provides fiscal data and served as inspiration to this endeavour.

# {.tabset .tabset-pills}

## Economy

\ 

```{r, echo = FALSE, out.width='100%', out.height='100%'}
fig_sales_serv + geom_line() + geom_point() + xlab('') + ylab('') + 
               scale_x_date(labels = scales::label_date(format = "%m"), date_breaks = "months", 
               expand = expansion(c(0.05, 0.05)), limits = c(min(serv_current$time), as.Date('2020-12-01'))) +
               scale_y_continuous(breaks = scales::breaks_width(20))+
               scale_colour_fira(name = element_text("")) + theme_fira() + theme(plot.margin = margin(0,5,10,-5), legend.title = element_text(face='bold'), plot.title = element_text(hjust = 0), plot.subtitle = element_text(hjust=0),
                                                                                 plot.caption = element_text(hjust=0)) +
               facet_wrap(~plot) + labs(title='Sales in services', subtitle='Monthly index, base = same period previous year', caption = paste0('https://vdb.czso.cz/vdbvo2/faces/index.jsf?page=vystup-objekt-parametry&pvo=SLU03-A&sp=A&skupId=1877&pvokc=&katalog=32025&z=T\n','last checked: ', format(date, '%d-%m-%Y')))
```

\ 

```{r, echo = FALSE, out.width='100%', out.height='100%'}
fig_prod_ind + geom_line() + geom_point() + xlab('') + ylab('') + 
               scale_x_date(labels = scales::label_date(format = "%m"), date_breaks = "months", 
               expand = expansion(c(0.05, 0.05)), limits = c(min(ind_current$time), as.Date('2020-12-01'))) +
               scale_colour_fira(name = element_text("")) + theme_fira() + theme(plot.margin = margin(0,5,10,-5), legend.title = element_text(face='bold'), plot.title = element_text(hjust = 0), plot.subtitle = element_text(hjust=0), plot.caption = element_text(hjust=0)) +
               facet_wrap(~categ_factor) + labs(title='Industry production', subtitle='Monthly index, base = same period previous year', caption = paste0('https://vdb.czso.cz/vdbvo2/faces/index.jsf?page=vystup-objekt-parametry&pvo=PRU01-D&sp=A&skupId=1267&pvokc=&katalog=30835&z=T\n','last checked: ', format(date, '%d-%m-%Y')))
```               

\ 

```{r, echo = FALSE, out.width='100%', out.height='100%'}
fig_prod_cons + geom_line() + geom_point() + xlab('') + ylab('') + 
               scale_x_date(labels = scales::label_date(format = "%m"), date_breaks = "months", 
               expand = expansion(c(0.05, 0.05)), limits = c(min(ind_current$time), as.Date('2020-12-01'))) +
               scale_colour_fira(name = element_text("")) + theme_fira() + theme(plot.margin = margin(0,5,10,-5), legend.title = element_text(face='bold'), plot.title = element_text(hjust = 0), plot.subtitle = element_text(hjust=0), plot.caption = element_text(hjust=0)) +
               facet_wrap(~plot) + labs(title='Construction production', subtitle='Monthly index, base = same period previous year', caption = paste0('https://vdb.czso.cz/vdbvo2/faces/cs/index.jsf?page=vystup-objekt-parametry&pvo=STA01-D&sp=A&skupId=826&pvokc=&katalog=30836&z=T\n','last checked: ', format(date, '%d-%m-%Y')))
```

\ 

```{r, echo = FALSE, out.width='100%', out.height='100%'}
plot1 <- fig_urate + geom_line() + geom_point()
plot2 <- fig_unem + geom_line() + geom_point()
grid.arrange(plot1, plot2, ncol=2)
```

\ 

```{r, echo = FALSE, out.width='100%', out.height='100%'}
grid.arrange(fig_infl, fig_infl_ppi, ncol=2)
```

\

```{r, echo = FALSE, out.width='100%', out.height='100%'}
grid.arrange(fig_loan_growth, fig_npls, ncol=2)
```

```{r, echo = FALSE, out.width='100%', out.height='100%'}
fig_morat
```

## Leading indicators

## Health

```{r mort-data}
zmr0 <- czso_get_table("130185", force_redownload = T)
zmr <- zmr0 %>% 
  mutate(tyden = as.numeric(tyden),
         casref_do = as.Date(casref_do),
         day_sameyear = make_date(1970, month = month(casref_do),
                                  day = day(casref_do))) %>% 
  mutate(vek_txt = fct_relevel(vek_txt, "celkem") %>% 
           fct_recode(Total = "celkem"))

last_date_fmt <- max(zmr$casref_do)
today_fmt <- today()
```

```{r time-adjusting-load}
vek <- read_rds("vek.rds")
vek_ciselnik <- read_rds("vek_ciselnik.rds")
vek_prevodnik <- read_csv("vek-prevodnik.csv")
```


```{r time-adjusting-age}
vek_slct <- vek %>%  
  filter(vuzemi_cis == "97") %>% 
  filter(is.na(pohlavi_txt)) %>% 
  left_join(vek_ciselnik %>%
                    select(vek_kod = CHODNOTA, zkr = ZKRTEXT, 
                           txt = TEXT, matches("MIN|MAX"))) %>% 
  select(idhod, hodnota, vek_txt, zkr, txt, casref_do) %>% 
  left_join(vek_prevodnik) %>% 
  group_by(casref_do, vek_txt = vek_grp) %>% 
  summarise(pocet_obyv = sum(hodnota, na.rm = T)) %>% 
  mutate(rok = year(casref_do)) %>% 
  replace_na(list(vek_txt = "Total")) %>% 
  ungroup() %>% 
  select(-casref_do)

vek_2019 <- vek_slct %>% 
  filter(rok == 2019, !is.na(vek_txt)) %>% 
  rename(obyv_2019 = pocet_obyv) %>% 
  select(-rok)

vek_do2020 <- vek_slct %>% 
  bind_rows(vek_slct %>% 
              filter(rok == 2019) %>% 
              mutate(rok = 2020))
```


```{r time-adjusting-age-adjust}
zmr_normd <- zmr %>% 
  filter(vek_txt != "Total") %>% 
  select(hodnota, rok, tyden, roktyden, vek_txt,
         idhod, stapro_kod, casref_do) %>% 
  left_join(vek_do2020) %>% 
  mutate(per_100k = hodnota/pocet_obyv * 1e5) %>% 
  left_join(vek_2019) %>% 
  mutate(normd = per_100k * obyv_2019 / 1e5)

zmr_normd_with_sum <- zmr_normd %>% 
  bind_rows(zmr_normd %>% 
              group_by(rok, tyden, roktyden, casref_do) %>% 
              summarise(across(c(hodnota, obyv_2019, pocet_obyv, normd), sum)) %>% 
              mutate(per_100k = hodnota/pocet_obyv * 1e5,
                     vek_txt = "Total")) %>% 
  mutate(day_sameyear = make_date(1970, month(casref_do), day(casref_do)))
```


```{r time-adjusting-prep}
zmr_ribbon_data <- zmr_normd_with_sum %>% 
  mutate(is2020 = rok == 2020) %>% 
  mutate(day_sameyear_floored = floor_date(day_sameyear, "weeks")) %>% 
  group_by(is2020, vek_txt, tyden, day_sameyear_floored) %>% 
  summarise(min = min(normd, na.rm = T), 
            max = max(normd, na.rm = T), .groups = "drop") %>% 
  mutate(vek_txt = fct_relevel(vek_txt, "Total") %>% 
           fct_recode(`85+` = "85 a více"))
```


```{r mort-labels}
lbs_grey <- tribble(~vek_txt, ~day_sameyear_floored, ~min, ~txt,
               factor("Total"), as.Date("1970-02-01"), 2650, "range in \n2011-19",
               )
lbs_blue <- tribble(~vek_txt, ~day_sameyear_floored, ~min, ~txt,
               factor("Total"), as.Date("1970-05-05"), 1900, "2020"
               )
```


```{r age-adjusted-plot}
ggplot(data = zmr_ribbon_data, mapping = aes(group = vek_txt)) +
  geom_ribbon(data = zmr_ribbon_data %>% filter(!is2020),
              alpha = 0.4,
              aes(x = day_sameyear_floored, ymin = min, ymax = max)) +
  geom_line(data = zmr_ribbon_data %>% filter(is2020),
            colour = "darkblue",
            aes(x = day_sameyear_floored, y = min)) +
  geom_point(data = zmr_ribbon_data %>% filter(is2020),
            aes(x = day_sameyear_floored, y = min), 
            size = 1, colour = "darkblue", fill = "white", shape = 21) +
  facet_wrap(~vek_txt, scales = "free_y") +
  scale_x_date(date_breaks = "3 months",
               date_labels = "%b", breaks = c(1,3,6, 9, 12)) +
  theme_fira() +
  theme(plot.margin = margin(0,5,10,0), 
        legend.title = element_text(face='bold'), 
        plot.title = element_text(hjust = 0), 
        plot.subtitle = element_text(hjust=0),
        plot.caption = element_text(hjust=0)) +
  labs(x = NULL, y = NULL,
       title = "Mortality (all-cause)",
       subtitle = "Deaths per week, 2011-2020. Blue = 2020, grey = range in 2011-2019",
       caption = str_glue("Based on CZSO dataset 130185 (deaths) and 130142 (age) | last checked {today_fmt} | Last data for {last_date_fmt}.")) +
  geom_text(data = lbs_grey, aes(x = day_sameyear_floored, 
                            y = min, label = txt), 
            family = "Fira Sans",
            fontface = "bold",
            colour = "white",size = 3) +
  geom_text(data = lbs_blue, aes(x = day_sameyear_floored, 
                            y = min, label = txt), 
            family = "Fira Sans",
            fontface = "bold",
            colour = "darkblue", size = 3)
```

<!-- ## Energy consumption 
#
## PX index
# -->

<!-- ## Some other useful indicators? -->