---
title: ""
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))}) 
author: "Jan Mareš"
# date: ""
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r back, echo = FALSE, include = FALSE}
library(data.table)
library(here)
library(czso)
library(dplyr)
library(stringr)
library(ggplot2)
library(eurostat)
library(readr)
library(firatheme)

##########################################################
# span of data, get the current year
date <- Sys.Date()
year_sel <- format(date, "%Y")

##########################################################
# unemployment
unemp <- get_eurostat('ei_lmhr_m')

unemp_cz <- as.data.table(unemp)
unemp_cz <- unemp_cz[geo == "CZ" & year(time) >= (as.numeric(year_sel)-1) &  s_adj == "SA" & indic == "LM-UN-T-TOT", ]

fig_u <- ggplot(unemp_cz, aes(x = time, y = values, colour = factor(1))) + xlab('') + ylab('') +
               scale_x_date(labels = scales::label_date(format = "%m"), date_breaks = "months", 
               expand = expansion(c(0.01, 0.01)), limits = c(min(unemp_cz$time), as.Date('2020-12-01'))) +         
               scale_y_continuous(limits = c(1.5,5)) +
               scale_colour_fira(name = element_text("")) + theme_fira() + 
               theme(legend.position = "none", legend.title = element_text(face='bold'), plot.title = element_text(hjust = 0), 
               plot.subtitle = element_text(hjust=0)) + 
               labs(title='Unemployment', subtitle='Harmonized rate of unemployment (Eurostat)')
#
# Podíl nezaměstnaných osob (MPSV - úřady práce)
# https://www.mpsv.cz/o/rest/statistiky/nezamestnanost/2020/05


```

## Macroeconomics dashboard - Czech Republic

This dashboard aims to provide basic macroeconomics data for the Czech Republic. The idea is to gather an up-to-date value from different sources and provide simple overview. The initial idea comes from the lack of unified information source about the economic development during the pandemic crises. Czech also [czechtrack](https://petrbouchal.github.io/czechtrack/) which provides fiscal data and served as inspiration to the current project.

```{r pokus, echo=FALSE}
fig_u + geom_line() + geom_point()
```
